---
title: "Experimental Design: Mining on Preassembled Genomes"
subtitle: "Analysis of gastric-cancer related biosynthetic gene clusters"
author: "LIMA, Carlos; SILVA JUNIOR, Wilson"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    highlight-style: tango
    df-print: paged
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5
)

# Load packages
library(tidyverse)

# Load data
data_path <- file.path("..", "copied_from_funcscan_results", "combgc_complete_summary.tsv")
if (!file.exists(data_path)) {
  stop(sprintf("Required data file not found: %s", data_path))
}
bgc_table <- read.csv(data_path,
                      header = TRUE,
                      sep = "\t")
```

# Overview

We downloaded 554 preassembled *H. pylori* genomes from the NCBI database, 179 of which were from gastric-cancer related samples.
The other 375 were from precancerous cases, such as chronic gastritis and intestinal metaplasia.

# Provisional Objectives

1. Agrupar os BGCs com a técnica de Gene Cluster Families (GCFs);
2. Correlacionar filogenia com repertórios de GCFs;
3. Comparar o repertórios de GCFs/BGCs entre os genomas categorizados em pré-malignos e câncer;
4. Identificar as GCFs estatisticamente relacionadas com câncer;
5. Analisar a relevância biológica do GCFs/BGCs com a progressão tumoral. Ou seja, identificar uma assinatura de malignidade
baseada nos tipos de GCFs/BGCs;
6. Por fim, discutir uma abordagem experimental para validação funcional dos GCFs/BGCs em modelo animal.

# Methods

## Genome mining and BGC profiling

We used the *nf-core* pipeline **funcscan** (version 3.0.0, running on Nextflow version
25.10.4) to perform genome mining on the preassembled genomes. BGC profiling was
performed with antiSMASH (version 8.0.1), deepBGC (version 0.1.31), and GECCO
(version 0.9.10). The mining tools' results were aggregated into a single table using
comBGC, as follows:

```{r display_bgc_table, echo=FALSE}
library(kableExtra)

knitr::kable(head(bgc_table, 20L)) |>
  kable_styling() |>
  scroll_box(width = "100%")
```

## BGC clustering into GCFs (adapted from Robey et al., 2021)

GCF clustering followed an approach adapted from Robey et al. (2021). Since BiG-SCAPE
requires antiSMASH-formatted GenBank files and is therefore incompatible with deepBGC
and GECCO outputs, we implemented a tool-agnostic workflow operating directly on the
aggregated comBGC summary table, which provides harmonized Pfam domain arrays for all
BGCs regardless of the prediction tool.

### BGC inventory across tools

The combined dataset comprised `r nrow(bgc_table)` BGC predictions across 554 genomes.
antiSMASH predicted three product classes: Terpene-precursor (n = 554, one locus per
genome), Azole-containing-RiPP (n = 19), and Arylpolyene (n = 1). deepBGC predictions
spanned Polyketide (n = 134), Saccharide (n = 125), Other (n = 83), RiPP (n = 61),
and NRP (n = 50), alongside 1,007 BGCs for which the model assigned high detection
probability (mean = 0.59, max = 0.92) but no product class — likely reflecting the
atypical secondary metabolome of *H. pylori* that falls outside the product classes
represented in the deepBGC training data. GECCO predicted 36 Unknown and 1 RiPP BGCs.

### Pre-clustering filters

Prior to similarity computation, the following filters were applied:

1. **deepBGC probability threshold**: deepBGC BGCs with `BGC_probability` < 0.50 were
   excluded, retaining only predictions at or above the tool's recommended detection
   threshold (Hannigan et al., 2019).

2. **Unclassified deepBGC BGCs**: the deepBGC predictions without a product class
   assignment were retained as a separate `Unknown` class and processed using Jaccard
   similarity only (see below).

3. **GECCO Unknown BGCs**: retained and processed as `Unknown` class alongside the
   unclassified deepBGC predictions.

### Domain annotation

Pfam domain arrays for each BGC were extracted from the `PFAM_domains` column of the
comBGC summary. For BGCs predicted by antiSMASH, domain annotations were derived from
hmmsearch results against the Pfam database as part of the funcscan/antiSMASH workflow.
For deepBGC and GECCO predictions, Pfam domains were assigned during the annotation
step executed upstream by funcscan.

### Pairwise similarity scoring

For each pair of BGCs, Pfam domain arrays were compared using Jaccard similarity,
computed with the Python package `scikit-learn`. Pairs with Jaccard similarity below
0.1 were excluded from further analysis.

Given the restricted biosynthetic repertoire of *H. pylori* — dominated by
Terpene-precursor, RiPP, and Arylpolyene clusters with limited canonical PKS or NRPS
backbone architecture — two similarity regimes were applied:

- **BGCs with identifiable backbone domains** (terpene synthases: PT_FPPS_like,
  PF02353; NRPS condensation-like: PF01135): the full Robey et al. score was applied,

$$\text{Similarity Score} = \sqrt{0.8 \times \text{sequence identity} + 0.2 \times \text{Jaccard similarity}}$$

  where sequence identity between backbone domains was estimated via `hmmalign` profile
  alignments. When multiple copies of a domain were present in a pair, the Hungarian
  matching algorithm was used to identify the highest-similarity domain pairing
  configuration, following Robey et al. (2021).

- **BGCs without backbone domains** (Unknown class and RiPP/Arylpolyene BGCs lacking
  canonical backbone enzymes): similarity was computed from Jaccard distance alone,

$$\text{Similarity Score} = \sqrt{0.2 \times \text{Jaccard similarity}}$$

### GCF delineation

DBSCAN clustering (epsilon = 0.50, min_samples = 2) was applied to the pairwise
distance matrix (distance = 1 − Similarity Score). BGCs that did not meet the minimum
neighborhood density criterion (DBSCAN label = −1) were designated as singletons and
not assigned to any GCF.

```{r gcf_clustering, echo=FALSE}
# TODO: load GCF clustering output
# gcf_assignments <- read.csv("path/to/gcf_assignments.tsv", sep = "\t")
```
