---
title: "GCF Clustering Results — Full Cohort Statistical Analysis"
subtitle: "Exploratory analysis of biosynthetic gene cluster families in *H. pylori*"
author: "LIMA, Carlos; SILVA JUNIOR, Wilson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)

library(tidyverse)
library(kableExtra)

# --- paths (relative to analysis/R/RMarkdown) ---
RESULTS_DIR <- file.path("..", "..", "scripts", "results")
METADATA    <- file.path("..", "..", "..", "rawdata", "metadata", "gbk_hp_genomes.tsv")
COMBGC      <- file.path("..", "copied_from_funcscan_results", "combgc_complete_summary.tsv")

# check required files exist
required_files <- c(
  file.path(RESULTS_DIR, "gcf_assignments.tsv"),
  file.path(RESULTS_DIR, "bgc_domain_arrays.tsv"),
  COMBGC,
  METADATA
)
missing_files <- required_files[!file.exists(required_files)]
if (length(missing_files) > 0) {
  stop(sprintf("Required data file(s) not found: %s", paste(missing_files, collapse = ", ")))
}

# --- load data ---
gcf     <- read_tsv(file.path(RESULTS_DIR, "gcf_assignments.tsv"),  show_col_types = FALSE)
domains <- read_tsv(file.path(RESULTS_DIR, "bgc_domain_arrays.tsv"), show_col_types = FALSE)
combgc  <- read_tsv(COMBGC, show_col_types = FALSE)
meta    <- read_tsv(METADATA, show_col_types = FALSE) |>
  rename(sample_id = Genome_ID) |>
  mutate(Phenotype = str_trim(Phenotype),
         group     = if_else(Phenotype == "Gastric cancer", "Cancer", "Non-cancer"))

# --- join GCF assignments with metadata ---
gcf_meta <- gcf |>
  left_join(meta, by = "sample_id")

# --- convenience flags ---
is_singleton <- gcf_meta$gcf_id == "singleton"

# --- harmonised product class ---
harmonise_product <- function(x) {
  case_when(
    str_detect(x, regex("terpene", ignore_case = TRUE)) ~ "Terpene",
    str_detect(x, regex("polyketide", ignore_case = TRUE)) ~ "Polyketide",
    str_detect(x, regex("NRP|NRPS|nrp", ignore_case = FALSE)) ~ "NRP",
    str_detect(x, regex("RiPP|ripp", ignore_case = TRUE)) ~ "RiPP",
    str_detect(x, regex("saccharide", ignore_case = TRUE)) ~ "Saccharide",
    str_detect(x, regex("arylpolyene", ignore_case = TRUE)) ~ "Arylpolyene",
    str_detect(x, regex("azole", ignore_case = TRUE)) ~ "Azole-RiPP",
    x == "" | is.na(x) ~ "Unclassified",
    TRUE ~ "Other"
  )
}

gcf_meta <- gcf_meta |>
  mutate(product_harmonised = harmonise_product(Product_class))
```

# Scope of this document

This document supersedes the statistical-testing sections of `gcf_results.Rmd`
without replacing it. The original document selected the 20 GCFs with the
**largest prevalence difference** between Cancer and Non-cancer genomes and
then tested those same 20 GCFs for significance. That procedure introduces a
selection bias: prevalence differences are noisiest for rare, small-sample
GCFs, so ranking by the observed difference and then testing preferentially
surfaces GCFs whose apparent effect is inflated by sampling variation
("winner's curse"). Multiple-testing correction applied only to that
pre-selected subset is therefore not a valid correction for the number of
GCFs actually considered during discovery.

Here, Fisher's exact test, the Cochran–Mantel–Haenszel (CMH) test, and the
multivariable logistic regression are all run over **every GCF that meets a
minimum-prevalence criterion** (≥ 3 genomes total, decided *a priori* and 
independent of any observed effect size), and Benjamini–Hochberg correction 
is applied across that full qualifying set. The "top 20 by prevalence 
difference" view is retained purely as a **descriptive** visualization — 
it is explicitly labelled as such and is not used to select which GCFs are 
tested.

# Overview

```{r overview-table}
n_raw <- nrow(combgc)

tibble(
  Metric = c(
    "Total BGCs predicted (pre-filter)",
    "Total BGCs (post-filter)",
    "BGCs assigned to a GCF",
    "Singletons",
    "Total GCFs",
    "Largest GCF (BGCs)",
    "Median GCF size",
    "Genomes with metadata"
  ),
  Value = c(
    n_raw,
    nrow(gcf_meta),
    sum(!is_singleton),
    sum(is_singleton),
    n_distinct(gcf_meta$gcf_id[!is_singleton]),
    gcf_meta |> filter(!is_singleton) |> count(gcf_id) |> pull(n) |> max(),
    gcf_meta |> filter(!is_singleton) |> count(gcf_id) |> pull(n) |> median(),
    sum(!is.na(gcf_meta$Phenotype))
  )
) |>
  knitr::kable() |>
  kable_styling(full_width = FALSE)
```

# BGC inventory

## Predictions per tool

```{r tool-barplot}
gcf_meta |>
  count(Prediction_tool, gcf_id) |>
  mutate(assigned = if_else(gcf_id == "singleton", "Singleton", "In GCF")) |>
  count(Prediction_tool, assigned) |>
  ggplot(aes(x = Prediction_tool, y = n, fill = assigned)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("In GCF" = "#2c7bb6", "Singleton" = "#d7191c")) +
  labs(x = NULL, y = "BGC count", fill = NULL,
       title = "BGC assignment by prediction tool") +
  theme_minimal(base_size = 13)
```

## Product class distribution

```{r product-class}
gcf_meta |>
  filter(product_harmonised != "Unclassified") |>
  count(Prediction_tool, product_harmonised) |>
  ggplot(aes(x = reorder(product_harmonised, n, sum),
             y = n,
             fill = Prediction_tool)) +
  geom_col(position = "stack") +
  coord_flip() +
  labs(x = NULL, y = "BGC count", fill = "Tool",
       title = "Product class distribution (harmonised)") +
  theme_minimal(base_size = 13)
```

# GCF characterization

## GCF size distribution

```{r gcf-size-dist}
gcf_sizes <- gcf_meta |>
  filter(!is_singleton) |>
  count(gcf_id, name = "size")

gcf_sizes |>
  filter(size <= 50) |>
  ggplot(aes(x = size)) +
  geom_histogram(binwidth = 1, fill = "#2c7bb6", color = "white") +
  labs(x = "GCF size (BGCs)", y = "Number of GCFs",
       title = "GCF size distribution (≤ 50 BGCs)",
       subtitle = sprintf("Dominant Terpene-precursor GCF (n = %d) excluded for legibility",
                          max(gcf_sizes$size))) +
  theme_minimal(base_size = 13)
```

## Top 15 GCFs

```{r top-gcfs}
gcf_meta |>
  filter(!is_singleton) |>
  group_by(gcf_id) |>
  summarise(
    size              = n(),
    dominant_backbone = names(sort(table(backbone_class), decreasing = TRUE))[1],
    dominant_product  = names(sort(table(product_harmonised), decreasing = TRUE))[1],
    dominant_tool     = names(sort(table(Prediction_tool), decreasing = TRUE))[1],
    n_tools           = n_distinct(Prediction_tool),
    .groups           = "drop"
  ) |>
  arrange(desc(size)) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("GCF", "Size", "Dominant backbone",
                              "Dominant product", "Dominant tool", "Tools")) |>
  kable_styling(full_width = FALSE) |>
  scroll_box(width = "100%")
```

# Clinical context

## BGC richness per genome

Genomes from gastric cancer patients harbour significantly more BGCs per genome
than non-cancer genomes (Wilcoxon rank-sum test, p < 0.001), suggesting that
oncogenic *H. pylori* lineages may carry a more diverse biosynthetic repertoire.

```{r bgc-richness}
richness <- gcf_meta |>
  filter(!is.na(group)) |>
  count(sample_id, group, name = "n_bgcs")

wt <- wilcox.test(n_bgcs ~ group, data = richness)

richness |>
  ggplot(aes(x = group, y = n_bgcs, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.2) +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  annotate("text", x = 1.5, y = max(richness$n_bgcs) * 0.95,
           label = sprintf("W = %.0f\np = %.4f", wt$statistic, wt$p.value),
           size = 4) +
  labs(x = NULL, y = "BGCs per genome", fill = NULL,
       title = "BGC richness by clinical group") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

Cancer genomes have a median of `r median(richness$n_bgcs[richness$group == "Cancer"])`
BGCs per genome vs `r median(richness$n_bgcs[richness$group == "Non-cancer"])` for
non-cancer genomes (W = `r wt$statistic`, p = `r format(wt$p.value, digits = 3)`).

## BGC richness per genome — by prediction tool

To assess whether the increased BGC richness in cancer genomes is consistent across
prediction tools or driven by unclassified deepBGC predictions, the analysis was
stratified by tool.

```{r bgc-richness-by-tool}
richness_tool <- gcf_meta |>
  filter(!is.na(group)) |>
  count(sample_id, group, Prediction_tool, name = "n_bgcs")

richness_tool |>
  ggplot(aes(x = group, y = n_bgcs, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.0) +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  facet_wrap(~ Prediction_tool, scales = "free_y") +
  labs(x = NULL, y = "BGCs per genome", fill = NULL,
       title = "BGC richness by clinical group — stratified by prediction tool") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

The tool-stratified pattern reveals that the signal of higher BGC richness in 
cancer genomes is not uniform across prediction methods. For antiSMASH, the 
distributions for Cancer and Non-cancer samples are virtually identical (W = 
33898.5, p_adj = 0.387, ns)—an expected outcome, given that this tool typically 
detects only 1–2 BGCs per genome (predominantly the universal terpene-precursor 
cluster), leaving little room for inter-group variation. In contrast, deepBGC 
shows a robust and highly significant difference (W = 37563.0, p_adj = 2.61e-05, 
***), with cancer genomes exhibiting higher median values and greater dispersion 
in the number of BGCs per genome. This indicates that the overall signal of 
increased richness in cancer is largely driven by deepBGC and does not represent 
a consistent, generalized trend across all mining tools. GECCO could not be 
evaluated due to the extremely low frequency of predictions in this dataset. 
This result is methodologically significant and it underscores the need for 
caution when interpreting this finding as a biological pattern.

````{r bgc-richness-by-tool-tests, cache=TRUE}
tools <- unique(gcf_meta$Prediction_tool)

map_dfr(tools, function(tool) {
  sub <- richness_tool |> filter(Prediction_tool == tool)

  if (n_distinct(sub$group) < 2) {
    return(tibble(Tool = tool, W = NA_real_, p_value = NA_real_, p_adj_BH = NA_real_))
  }

  wt <- wilcox.test(n_bgcs ~ group, data = sub)
  tibble(
    Tool    = tool,
    W       = wt$statistic,
    p_value = wt$p.value,
    p_adj_BH = NA_real_
  )
}) |>
  mutate(p_adj_BH = p.adjust(p_value, method = "BH"),
         sig = case_when(
           p_adj_BH < 0.001 ~ "***",
           p_adj_BH < 0.01  ~ "**",
           p_adj_BH < 0.05  ~ "*",
           is.na(p_adj_BH)  ~ "—",
           TRUE             ~ "ns"
         ),
         p_value  = formatC(p_value,  format = "e", digits = 2),
         p_adj_BH = formatC(p_adj_BH, format = "e", digits = 2)) |>
  knitr::kable(col.names = c("Tool", "W", "p", "p_adj (BH)", "Sig.")) |>
  kable_styling(full_width = FALSE)
````

## GCF richness per genome

```{r gcf-richness}
gcf_richness <- gcf_meta |>
  filter(!is_singleton & !is.na(group)) |>
  distinct(sample_id, gcf_id, group) |>
  count(sample_id, group, name = "n_gcfs")

wt2 <- wilcox.test(n_gcfs ~ group, data = gcf_richness)

gcf_richness |>
  ggplot(aes(x = group, y = n_gcfs, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.2) +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  annotate("text", x = 1.5, y = max(gcf_richness$n_gcfs) * 0.95,
           label = sprintf("W = %.0f\np = %.4f", wt2$statistic, wt2$p.value),
           size = 4) +
  labs(x = NULL, y = "GCFs per genome", fill = NULL,
       title = "GCF richness by clinical group") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

Cancer genomes carry a median of `r median(gcf_richness$n_gcfs[gcf_richness$group == "Cancer"])`
distinct GCFs per genome vs `r median(gcf_richness$n_gcfs[gcf_richness$group == "Non-cancer"])`
for non-cancer (W = `r wt2$statistic`, p = `r format(wt2$p.value, digits = 3)`).

## GCF prevalence: Cancer vs Non-cancer — full cohort

For each non-singleton GCF with at least 3 carrier genomes in total (the
pre-specified minimum-size filter; see Scope above), prevalence was computed
as the fraction of genomes in each clinical group carrying at least one BGC
assigned to that family. Significance was assessed using Fisher's exact test.
P-values were adjusted for multiple testing using the Benjamini–Hochberg (FDR)
procedure **across the entire qualifying set**, not a pre-selected subset.

```{r gcf-prevalence-full, cache=TRUE}
MIN_GENOMES_TOTAL <- 3

n_cancer    <- sum(meta$group == "Cancer",     na.rm = TRUE)
n_noncancer <- sum(meta$group == "Non-cancer", na.rm = TRUE)

# presence/absence per genome per GCF
presence <- gcf_meta |>
  filter(!is_singleton & !is.na(group)) |>
  distinct(sample_id, gcf_id, group)

# a priori size filter: GCF must be carried by >= MIN_GENOMES_TOTAL genomes
gcf_carrier_counts <- presence |>
  count(gcf_id, name = "n_carrier_genomes")

qualifying_gcfs <- gcf_carrier_counts |>
  filter(n_carrier_genomes >= MIN_GENOMES_TOTAL) |>
  pull(gcf_id)

n_gcf_total      <- n_distinct(presence$gcf_id)
n_gcf_qualifying <- length(qualifying_gcfs)

# Fisher's exact test per GCF
fisher_per_gcf <- function(gcf_id_val) {
  sub <- presence |> filter(gcf_id == gcf_id_val)
  a <- sum(sub$group == "Cancer")
  b <- sum(sub$group == "Non-cancer")
  c <- n_cancer    - a
  d <- n_noncancer - b
  mat <- matrix(c(a, b, c, d), nrow = 2)
  ft  <- fisher.test(mat)
  tibble(gcf_id      = gcf_id_val,
         n_cancer    = a,
         n_noncancer = b,
         prev_cancer    = a / .env$n_cancer,
         prev_noncancer = b / .env$n_noncancer,
         odds_ratio  = ft$estimate,
         p_value     = ft$p.value)
}

gcf_stats_full <- map_dfr(qualifying_gcfs, fisher_per_gcf) |>
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    diff  = prev_cancer - prev_noncancer,
    sig   = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    )
  ) |>
  arrange(p_adj)
```

`r n_gcf_total` non-singleton GCFs were identified in total; `r n_gcf_qualifying`
met the ≥ `r MIN_GENOMES_TOTAL`-genome minimum-size criterion and were carried
forward into Fisher/CMH/GLM testing (the remaining GCFs are supported by too
few genomes for any of these tests to have meaningful power, and were excluded
from testing *before* looking at their prevalence difference).

```{r gcf-prevalence-full-summary}
tibble(
  Metric = c(
    "GCFs tested (Fisher, full cohort)",
    "GCFs with p_adj < 0.05",
    "GCFs with p_adj < 0.10",
    "Minimum p_adj observed"
  ),
  Value = c(
    n_gcf_qualifying,
    sum(gcf_stats_full$p_adj < 0.05),
    sum(gcf_stats_full$p_adj < 0.10),
    formatC(min(gcf_stats_full$p_adj), format = "e", digits = 2)
  )
) |>
  knitr::kable() |>
  kable_styling(full_width = FALSE)
```

### Descriptive view: top 20 by prevalence difference (illustrative only)

The panel below reproduces the descriptive ranking used in `gcf_results.Rmd`
for visual comparability. **It is not the basis for any significance claim**:
the GCFs shown here are simply the 20 with the largest observed prevalence
difference among the `r n_gcf_qualifying` GCFs that passed the a priori size
filter. Formal inference (Fisher/CMH/GLM, with BH correction over the full
qualifying set) is reported in the sections that follow.

```{r gcf-prevalence-illustrative}
top20_illustrative <- gcf_stats_full |>
  arrange(desc(abs(diff))) |>
  slice_head(n = 20)

top20_illustrative |>
  pivot_longer(cols = c(prev_cancer, prev_noncancer),
               names_to = "group", values_to = "prevalence") |>
  mutate(group = if_else(group == "prev_cancer", "Cancer", "Non-cancer")) |>
  ggplot(aes(x = reorder(gcf_id, abs(diff)), y = prevalence, fill = group)) +
  geom_col(position = "dodge") +
  geom_text(data = top20_illustrative,
            aes(x = gcf_id, y = pmax(prev_cancer, prev_noncancer) + 0.02,
                label = sig),
            inherit.aes = FALSE, size = 4) +
  coord_flip() +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Prevalence", fill = NULL,
       title = "Illustrative: top 20 GCFs by prevalence difference",
       subtitle = "Descriptive ranking only — not used to select GCFs for testing",
       caption = "Significance: *** p_adj < 0.001 | ** p_adj < 0.01 | * p_adj < 0.05 | ns\n(p_adj computed over the full qualifying set of GCFs, not just those shown here)") +
  theme_minimal(base_size = 12)
```

### Contingency table — top 20 by prevalence difference (illustrative)

```{r gcf-contingency-table-full}
contingency_table_full <- top20_illustrative |>
  dplyr::mutate(
    cancer_absent    = .env$n_cancer    - .data$n_cancer,
    noncancer_absent = .env$n_noncancer - .data$n_noncancer
  ) |>
  dplyr::select(gcf_id, n_cancer, cancer_absent, n_noncancer, noncancer_absent)

contingency_table_full |>
  knitr::kable(
    col.names = c("GCF", "Cancer (present)", "Cancer (absent)",
                  "Non-cancer (present)", "Non-cancer (absent)"),
    align = "lcccc"
  ) |>
  kable_styling(full_width = FALSE, font_size = 11) |>
  scroll_box(width = "100%")
```

### Full-cohort Fisher results — top 20 by adjusted p-value

This is the analytically relevant ranking: the qualifying GCFs sorted by
Fisher p_adj (BH, corrected across all `r n_gcf_qualifying` qualifying GCFs).

```{r gcf-fisher-ranked-full}
domains_gcf <- domains |>
  mutate(bgc_id = paste0(sample_id, "__", contig_id)) |>
  left_join(gcf |> select(bgc_id, gcf_id), by = "bgc_id")

top_domains_all <- domains_gcf |>
  filter(gcf_id %in% gcf_stats_full$gcf_id) |>
  separate_rows(domain_array, sep = "\\|") |>
  filter(domain_array != "") |>
  count(gcf_id, domain_array, sort = TRUE) |>
  group_by(gcf_id) |>
  slice_max(n, n = 3) |>
  summarise(top_domains = paste(domain_array, collapse = "; "), .groups = "drop")

gcf_char_all <- gcf_meta |>
  filter(gcf_id %in% gcf_stats_full$gcf_id) |>
  group_by(gcf_id) |>
  summarise(
    size     = n(),
    backbone = names(sort(table(backbone_class),     decreasing = TRUE))[1],
    product  = names(sort(table(product_harmonised), decreasing = TRUE))[1],
    .groups  = "drop"
  )

gcf_stats_full |>
  slice_head(n = 20) |>
  select(gcf_id, n_cancer, n_noncancer, prev_cancer, prev_noncancer,
         odds_ratio, p_value, p_adj, sig) |>
  left_join(gcf_char_all,    by = "gcf_id") |>
  left_join(top_domains_all, by = "gcf_id") |>
  mutate(
    prev_cancer    = scales::percent(prev_cancer,    accuracy = 0.1),
    prev_noncancer = scales::percent(prev_noncancer, accuracy = 0.1),
    odds_ratio     = round(odds_ratio, 2),
    p_value        = formatC(p_value,  format = "e", digits = 2),
    p_adj          = formatC(p_adj,    format = "e", digits = 2)
  ) |>
  knitr::kable(
    col.names = c("GCF", "n Cancer", "n Non-cancer",
                  "Prev. Cancer", "Prev. Non-cancer",
                  "OR", "p", "p_adj (BH)", "Sig.",
                  "Size", "Backbone", "Product", "Top domains")
  ) |>
  kable_styling(font_size = 11) |>
  scroll_box(width = "100%")
```

## Adjusting for potential confounders — full cohort

As in the prevalence analysis above, adjustment analyses (CMH and multivariable
logistic regression) were run over **all qualifying GCFs** (≥ 3 carrier
genomes), and Benjamini–Hochberg correction was applied across that full set.

```{r gcf_logistic_full, cache=TRUE}
# ---------------------------------------------------------------------------
# Covariates: continent, age, total BGC richness per genome (confounder control)
# ---------------------------------------------------------------------------
covars <- meta |>
  dplyr::select(sample_id, Continent, Age) |>
  dplyr::mutate(Age = as.numeric(Age))

group_lookup <- meta |>
  dplyr::select(sample_id, group)

richness_per_genome <- gcf_meta |>
  dplyr::count(sample_id, name = "n_bgcs_total")

presence_wide <- presence |>
  dplyr::left_join(covars, by = "sample_id") |>
  dplyr::left_join(richness_per_genome, by = "sample_id") |>
  dplyr::filter(!is.na(group), !is.na(Continent))

all_genomes <- group_lookup |>
  dplyr::left_join(covars, by = "sample_id") |>
  dplyr::left_join(richness_per_genome, by = "sample_id") |>
  dplyr::filter(!is.na(group), !is.na(Continent))

# ---------------------------------------------------------------------------
# 1. Cochran-Mantel-Haenszel, stratified by continent
# ---------------------------------------------------------------------------
cmh_per_gcf <- function(gcf_id_val) {
  carriers <- presence_wide |> dplyr::filter(gcf_id == gcf_id_val) |> dplyr::pull(sample_id)

  tab_df <- all_genomes |>
    dplyr::mutate(carrier = sample_id %in% carriers) |>
    dplyr::count(Continent, group, carrier)

  strata <- unique(tab_df$Continent)
  arr_list <- list()
  for (s in strata) {
    sub <- tab_df |> dplyr::filter(Continent == s)
    m <- matrix(0, nrow = 2, ncol = 2,
                dimnames = list(carrier = c("TRUE", "FALSE"),
                                 group   = c("Cancer", "Non-cancer")))
    for (i in seq_len(nrow(sub))) {
      m[as.character(sub$carrier[i]), sub$group[i]] <- sub$n[i]
    }
    if (all(rowSums(m) > 0) && all(colSums(m) > 0)) {
      arr_list[[s]] <- m
    }
  }

  if (length(arr_list) < 1) {
    return(tibble(gcf_id = gcf_id_val, cmh_p = NA_real_, cmh_statistic = NA_real_))
  }

  arr <- array(unlist(arr_list), dim = c(2, 2, length(arr_list)))
  res <- tryCatch(mantelhaen.test(arr, correct = TRUE), error = function(e) NULL)

  if (is.null(res)) {
    return(tibble(gcf_id = gcf_id_val, cmh_p = NA_real_, cmh_statistic = NA_real_))
  }

  tibble(gcf_id = gcf_id_val,
         cmh_p         = res$p.value,
         cmh_statistic = unname(res$statistic))
}

cmh_results_full <- purrr::map_dfr(gcf_stats_full$gcf_id, cmh_per_gcf)

# ---------------------------------------------------------------------------
# 2. Logistic regression: GCF_presence ~ phenotype + continent + age + richness
# ---------------------------------------------------------------------------
glm_per_gcf <- function(gcf_id_val) {
  carriers <- presence_wide |> dplyr::filter(gcf_id == gcf_id_val) |> dplyr::pull(sample_id)

  dat <- all_genomes |>
    dplyr::mutate(
      carrier = as.integer(sample_id %in% carriers),
      group   = factor(group, levels = c("Non-cancer", "Cancer")),
      Continent = factor(Continent)
    ) |>
    dplyr::filter(!is.na(Age))

  if (dplyr::n_distinct(dat$carrier) < 2 || nrow(dat) < 20) {
    return(tibble(gcf_id = gcf_id_val, glm_or = NA_real_,
                   glm_p = NA_real_, glm_ci_low = NA_real_, glm_ci_high = NA_real_))
  }

  fit <- tryCatch(
    glm(carrier ~ group + Continent + Age + n_bgcs_total,
        data = dat, family = binomial()),
    error = function(e) NULL,
    warning = function(w) suppressWarnings(
      glm(carrier ~ group + Continent + Age + n_bgcs_total,
          data = dat, family = binomial())
    )
  )

  if (is.null(fit) || !"groupCancer" %in% names(coef(fit))) {
    return(tibble(gcf_id = gcf_id_val, glm_or = NA_real_,
                   glm_p = NA_real_, glm_ci_low = NA_real_, glm_ci_high = NA_real_))
  }

  est <- summary(fit)$coefficients["groupCancer", ]
  ci  <- tryCatch(confint.default(fit)["groupCancer", ],
                   error = function(e) c(NA_real_, NA_real_))

  tibble(
    gcf_id      = gcf_id_val,
    glm_or      = exp(est["Estimate"]),
    glm_p       = est["Pr(>|z|)"],
    glm_ci_low  = exp(ci[1]),
    glm_ci_high = exp(ci[2])
  )
}

glm_results_full <- purrr::map_dfr(gcf_stats_full$gcf_id, glm_per_gcf)

# ---------------------------------------------------------------------------
# 3. Combined comparison table (full qualifying cohort)
# ---------------------------------------------------------------------------
comparison_table_full <- gcf_stats_full |>
  dplyr::select(gcf_id, n_cancer, n_noncancer, odds_ratio,
         fisher_p = p_value, fisher_p_adj = p_adj) |>
  dplyr::left_join(cmh_results_full, by = "gcf_id") |>
  dplyr::left_join(glm_results_full, by = "gcf_id") |>
  dplyr::mutate(
    cmh_p_adj = p.adjust(cmh_p, method = "BH"),
    glm_p_adj = p.adjust(glm_p, method = "BH")
  ) |>
  dplyr::arrange(glm_p_adj)

n_cmh_testable <- sum(!is.na(comparison_table_full$cmh_p))
n_glm_testable <- sum(!is.na(comparison_table_full$glm_p))
n_cmh_sig <- sum(comparison_table_full$cmh_p_adj < 0.05, na.rm = TRUE)
n_glm_sig <- sum(comparison_table_full$glm_p_adj < 0.05, na.rm = TRUE)
```

```{r gcf-adjustment-summary-full}
tibble(
  Metric = c(
    "GCFs entering CMH/GLM (qualifying set)",
    "GCFs with a computable CMH statistic",
    "GCFs with a computable GLM estimate",
    "GCFs significant by CMH (p_adj < 0.05)",
    "GCFs significant by GLM (p_adj < 0.05)"
  ),
  Value = c(
    n_gcf_qualifying,
    n_cmh_testable,
    n_glm_testable,
    n_cmh_sig,
    n_glm_sig
  )
) |>
  knitr::kable() |>
  kable_styling(full_width = FALSE)
```

### Full comparison table — top 20 by GLM adjusted p-value

```{r gcf-comparison-table-display-full}
comparison_table_full |>
  slice_head(n = 20) |>
  mutate(
    fisher_p_adj = formatC(as.numeric(fisher_p_adj), format = "e", digits = 2),
    cmh_p_adj = formatC(cmh_p_adj, format = "e", digits = 2),
    glm_p_adj = formatC(glm_p_adj, format = "e", digits = 2),
    odds_ratio = round(odds_ratio, 2),
    glm_or     = round(glm_or, 2),
    glm_ci     = sprintf("%.2f–%.2f", glm_ci_low, glm_ci_high)
  ) |>
  select(gcf_id, n_cancer, n_noncancer,
         odds_ratio, fisher_p_adj,
         cmh_p_adj,
         glm_or, glm_ci, glm_p_adj) |>
  knitr::kable(
    col.names = c("GCF", "n Cancer", "n Non-cancer",
                  "OR (Fisher)", "p_adj (Fisher)",
                  "p_adj (CMH, continent-stratified)",
                  "aOR (GLM)", "95% CI (aOR)", "p_adj (GLM)")
  ) |>
  kable_styling(font_size = 11) |>
  scroll_box(width = "100%")
```

Extending inference to the full set of `r n_gcf_qualifying` GCFs meeting the
minimum-size criterion, and correcting for multiple testing across that
complete set, `r n_cmh_sig` GCF(s) reached CMH significance (p_adj < 0.05)
and `r n_glm_sig` GCF(s) reached significance in the multivariable logistic
regression (p_adj < 0.05) after adjusting for continent, host age, and total
BGC richness per genome. This confirms, on a broader and pre-specified basis,
the conclusion already suggested by the top-20 analysis in `gcf_results.Rmd`:
no individual GCF shows a robust, confounder-adjusted association with
gastric cancer phenotype in this cohort once multiple testing is properly
accounted for across the full set of candidates. The near-significant signals
observed for individual GCFs in the illustrative top-20 view are consistent
with sampling variability in small carrier groups rather than a reproducible
phenotype-specific enrichment.

These results do not rule out a biological association between BGC/GCF
content and gastric cancer — they indicate that, with the current sample
size and GCF granularity, no single GCF's presence/absence carries a
statistically robust signal after appropriate correction. Aggregate or
richness-based signals (e.g., total BGC/GCF richness per genome, see above)
remain the more robust univariate findings in this dataset.

## Continent distribution

```{r continent}
meta |>
  count(Continent, group) |>
  ggplot(aes(x = Continent, y = n, fill = group)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  labs(x = NULL, y = "Genomes", fill = NULL,
       title = "Sample distribution by continent and clinical group") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Age distribution

```{r age}
meta |>
  filter(!is.na(Age)) |>
  mutate(Age = as.numeric(Age)) |>
  ggplot(aes(x = group, y = Age, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.2) +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  labs(x = NULL, y = "Age (years)", fill = NULL,
       title = "Age distribution by clinical group") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```
