---
title: "GCF Clustering Results"
subtitle: "Exploratory analysis of biosynthetic gene cluster families in *H. pylori*"
author: "LIMA, Carlos; SILVA JUNIOR, Wilson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)

library(tidyverse)
library(kableExtra)

# --- paths ---
BASE_DIR    <- "/home/clima/projects/bgcs_in_cancer/mt/mining_on_preassembled_genomes"
RESULTS_DIR <- file.path(BASE_DIR, "analysis/scripts/results")
METADATA    <- file.path(BASE_DIR, "rawdata/metadata/gbk_hp_genomes.tsv")
COMBGC      <- file.path(BASE_DIR, "analysis/R/copied_from_funcscan_results/combgc_complete_summary.tsv")

# --- load data ---
gcf     <- read_tsv(file.path(RESULTS_DIR, "gcf_assignments.tsv"),  show_col_types = FALSE)
domains <- read_tsv(file.path(RESULTS_DIR, "bgc_domain_arrays.tsv"), show_col_types = FALSE)
combgc  <- read_tsv(COMBGC, show_col_types = FALSE)
meta    <- read_tsv(METADATA, show_col_types = FALSE) |>
  rename(sample_id = Genome_ID) |>
  mutate(Phenotype = str_trim(Phenotype),
         group     = if_else(Phenotype == "Gastric cancer", "Cancer", "Non-cancer"))

# --- join GCF assignments with metadata ---
gcf_meta <- gcf |>
  left_join(meta, by = "sample_id")

# --- convenience flags ---
is_singleton <- gcf_meta$gcf_id == "singleton"

# --- harmonised product class ---
harmonise_product <- function(x) {
  case_when(
    str_detect(x, regex("terpene", ignore_case = TRUE)) ~ "Terpene",
    str_detect(x, regex("polyketide", ignore_case = TRUE)) ~ "Polyketide",
    str_detect(x, regex("NRP|NRPS|nrp", ignore_case = FALSE)) ~ "NRP",
    str_detect(x, regex("RiPP|ripp", ignore_case = TRUE)) ~ "RiPP",
    str_detect(x, regex("saccharide", ignore_case = TRUE)) ~ "Saccharide",
    str_detect(x, regex("arylpolyene", ignore_case = TRUE)) ~ "Arylpolyene",
    str_detect(x, regex("azole", ignore_case = TRUE)) ~ "Azole-RiPP",
    x == "" | is.na(x) ~ "Unclassified",
    TRUE ~ "Other"
  )
}

gcf_meta <- gcf_meta |>
  mutate(product_harmonised = harmonise_product(Product_class))
```

# Overview

```{r overview-table}
n_raw <- nrow(combgc)

tibble(
  Metric = c(
    "Total BGCs predicted (pre-filter)",
    "Total BGCs (post-filter)",
    "BGCs assigned to a GCF",
    "Singletons",
    "Total GCFs",
    "Largest GCF (BGCs)",
    "Median GCF size",
    "Genomes with metadata"
  ),
  Value = c(
    n_raw,
    nrow(gcf_meta),
    sum(!is_singleton),
    sum(is_singleton),
    n_distinct(gcf_meta$gcf_id[!is_singleton]),
    gcf_meta |> filter(!is_singleton) |> count(gcf_id) |> pull(n) |> max(),
    gcf_meta |> filter(!is_singleton) |> count(gcf_id) |> pull(n) |> median(),
    sum(!is.na(gcf_meta$Phenotype))
  )
) |>
  knitr::kable() |>
  kable_styling(full_width = FALSE)
```

# BGC inventory

## Predictions per tool

```{r tool-barplot}
gcf_meta |>
  count(Prediction_tool, gcf_id) |>
  mutate(assigned = if_else(gcf_id == "singleton", "Singleton", "In GCF")) |>
  count(Prediction_tool, assigned) |>
  ggplot(aes(x = Prediction_tool, y = n, fill = assigned)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("In GCF" = "#2c7bb6", "Singleton" = "#d7191c")) +
  labs(x = NULL, y = "BGC count", fill = NULL,
       title = "BGC assignment by prediction tool") +
  theme_minimal(base_size = 13)
```

## Product class distribution

```{r product-class}
gcf_meta |>
  filter(product_harmonised != "Unclassified") |>
  count(Prediction_tool, product_harmonised) |>
  ggplot(aes(x = reorder(product_harmonised, n, sum),
             y = n,
             fill = Prediction_tool)) +
  geom_col(position = "stack") +
  coord_flip() +
  labs(x = NULL, y = "BGC count", fill = "Tool",
       title = "Product class distribution (harmonised)") +
  theme_minimal(base_size = 13)
```

# GCF characterization

## GCF size distribution

```{r gcf-size-dist}
gcf_sizes <- gcf_meta |>
  filter(!is_singleton) |>
  count(gcf_id, name = "size")

gcf_sizes |>
  filter(size <= 50) |>
  ggplot(aes(x = size)) +
  geom_histogram(binwidth = 1, fill = "#2c7bb6", color = "white") +
  labs(x = "GCF size (BGCs)", y = "Number of GCFs",
       title = "GCF size distribution (≤ 50 BGCs)",
       subtitle = sprintf("Dominant Terpene-precursor GCF (n = %d) excluded for legibility",
                          max(gcf_sizes$size))) +
  theme_minimal(base_size = 13)
```

## Top 15 GCFs

```{r top-gcfs}
gcf_meta |>
  filter(!is_singleton) |>
  group_by(gcf_id) |>
  summarise(
    size              = n(),
    dominant_backbone = names(sort(table(backbone_class), decreasing = TRUE))[1],
    dominant_product  = names(sort(table(product_harmonised), decreasing = TRUE))[1],
    dominant_tool     = names(sort(table(Prediction_tool), decreasing = TRUE))[1],
    n_tools           = n_distinct(Prediction_tool),
    .groups           = "drop"
  ) |>
  arrange(desc(size)) |>
  slice_head(n = 15) |>
  knitr::kable(col.names = c("GCF", "Size", "Dominant backbone",
                              "Dominant product", "Dominant tool", "Tools")) |>
  kable_styling(full_width = FALSE) |>
  scroll_box(width = "100%")
```

# Clinical context

## BGC richness per genome

Genomes from gastric cancer patients harbour significantly more BGCs per genome
than non-cancer genomes (Wilcoxon rank-sum test, p < 0.001), suggesting that
oncogenic *H. pylori* lineages may carry a more diverse biosynthetic repertoire.

```{r bgc-richness}
richness <- gcf_meta |>
  filter(!is.na(group)) |>
  count(sample_id, group, name = "n_bgcs")

wt <- wilcox.test(n_bgcs ~ group, data = richness)

richness |>
  ggplot(aes(x = group, y = n_bgcs, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.2) +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  annotate("text", x = 1.5, y = max(richness$n_bgcs) * 0.95,
           label = sprintf("W = %.0f\np = %.4f", wt$statistic, wt$p.value),
           size = 4) +
  labs(x = NULL, y = "BGCs per genome", fill = NULL,
       title = "BGC richness by clinical group") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

Cancer genomes have a median of `r median(richness$n_bgcs[richness$group == "Cancer"])`
BGCs per genome vs `r median(richness$n_bgcs[richness$group == "Non-cancer"])` for
non-cancer genomes (W = `r wt$statistic`, p = `r format(wt$p.value, digits = 3)`).

## BGC richness per genome — by prediction tool

To assess whether the increased BGC richness in cancer genomes is consistent across
prediction tools or driven by unclassified deepBGC predictions, the analysis was
stratified by tool.

```{r bgc-richness-by-tool}
richness_tool <- gcf_meta |>
  filter(!is.na(group)) |>
  count(sample_id, group, Prediction_tool, name = "n_bgcs")

richness_tool |>
  ggplot(aes(x = group, y = n_bgcs, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.0) +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  facet_wrap(~ Prediction_tool, scales = "free_y") +
  labs(x = NULL, y = "BGCs per genome", fill = NULL,
       title = "BGC richness by clinical group — stratified by prediction tool") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

````{r bgc-richness-by-tool-tests, cache=TRUE}
tools <- unique(gcf_meta$Prediction_tool)

map_dfr(tools, function(tool) {
  sub <- richness_tool |> filter(Prediction_tool == tool)

  # skip if fewer than 2 groups represented
  if (n_distinct(sub$group) < 2) {
    return(tibble(Tool = tool, W = NA_real_, p_value = NA_real_, p_adj_BH = NA_real_))
  }

  wt <- wilcox.test(n_bgcs ~ group, data = sub)
  tibble(
    Tool    = tool,
    W       = wt$statistic,
    p_value = wt$p.value,
    p_adj_BH = NA_real_
  )
}) |>
  mutate(p_adj_BH = p.adjust(p_value, method = "BH"),
         sig = case_when(
           p_adj_BH < 0.001 ~ "***",
           p_adj_BH < 0.01  ~ "**",
           p_adj_BH < 0.05  ~ "*",
           is.na(p_adj_BH)  ~ "—",
           TRUE             ~ "ns"
         ),
         p_value  = formatC(p_value,  format = "e", digits = 2),
         p_adj_BH = formatC(p_adj_BH, format = "e", digits = 2)) |>
  knitr::kable(col.names = c("Tool", "W", "p", "p_adj (BH)", "Sig.")) |>
  kable_styling(full_width = FALSE)
````

## GCF richness per genome

```{r gcf-richness}
gcf_richness <- gcf_meta |>
  filter(!is_singleton & !is.na(group)) |>
  distinct(sample_id, gcf_id, group) |>
  count(sample_id, group, name = "n_gcfs")

wt2 <- wilcox.test(n_gcfs ~ group, data = gcf_richness)

gcf_richness |>
  ggplot(aes(x = group, y = n_gcfs, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.2) +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  annotate("text", x = 1.5, y = max(gcf_richness$n_gcfs) * 0.95,
           label = sprintf("W = %.0f\np = %.4f", wt2$statistic, wt2$p.value),
           size = 4) +
  labs(x = NULL, y = "GCFs per genome", fill = NULL,
       title = "GCF richness by clinical group") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

Cancer genomes carry a median of `r median(gcf_richness$n_gcfs[gcf_richness$group == "Cancer"])`
distinct GCFs per genome vs `r median(gcf_richness$n_gcfs[gcf_richness$group == "Non-cancer"])`
for non-cancer (W = `r wt2$statistic`, p = `r format(wt2$p.value, digits = 3)`).

## GCF prevalence: Cancer vs Non-cancer

For each GCF, prevalence was computed as the fraction of genomes in each clinical group
carrying at least one BGC assigned to that family. Significance was assessed using
Fisher's exact test, which handles zero-count cells robustly. P-values were
adjusted for multiple testing using the Benjamini-Hochberg (FDR) procedure.
GCFs present exclusively in one group may represent lineage-specific biosynthetic
signatures of oncogenic *H. pylori* clones.

```{r gcf-prevalence, cache=TRUE}
n_cancer    <- sum(meta$group == "Cancer",     na.rm = TRUE)
n_noncancer <- sum(meta$group == "Non-cancer", na.rm = TRUE)

# presence/absence per genome per GCF
presence <- gcf_meta |>
  filter(!is_singleton & !is.na(group)) |>
  distinct(sample_id, gcf_id, group)

# Fisher's exact test per GCF
fisher_per_gcf <- function(gcf_id_val) {
  sub <- presence |> filter(gcf_id == gcf_id_val)
  a <- sum(sub$group == "Cancer")
  b <- sum(sub$group == "Non-cancer")
  c <- n_cancer    - a
  d <- n_noncancer - b
  mat <- matrix(c(a, b, c, d), nrow = 2)
  ft  <- fisher.test(mat)
  tibble(gcf_id      = gcf_id_val,
         n_cancer    = a,
         n_noncancer = b,
         prev_cancer    = a / n_cancer,
         prev_noncancer = b / n_noncancer,
         odds_ratio  = ft$estimate,
         p_value     = ft$p.value)
}

gcf_stats <- map_dfr(unique(presence$gcf_id), fisher_per_gcf) |>
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    diff  = prev_cancer - prev_noncancer,
    sig   = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ "ns"
    )
  ) |>
  arrange(desc(abs(diff)))

top20 <- gcf_stats |> slice_head(n = 20)

top20 |>
  pivot_longer(cols = c(prev_cancer, prev_noncancer),
               names_to = "group", values_to = "prevalence") |>
  mutate(group = if_else(group == "prev_cancer", "Cancer", "Non-cancer"),
         label = if_else(group == "Cancer",
                         paste0(gcf_id, " ", sig), gcf_id)) |>
  ggplot(aes(x = reorder(gcf_id, abs(diff)), y = prevalence, fill = group)) +
  geom_col(position = "dodge") +
  geom_text(data = top20,
            aes(x = gcf_id, y = pmax(prev_cancer, prev_noncancer) + 0.02,
                label = sig),
            inherit.aes = FALSE, size = 4) +
  coord_flip() +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = NULL, y = "Prevalence", fill = NULL,
       title = "Top 20 GCFs by prevalence difference (Cancer vs Non-cancer)",
       caption = "Significance: *** p_adj < 0.001 | ** p_adj < 0.01 | * p_adj < 0.05 | ns") +
  theme_minimal(base_size = 12)
```

### GCF details

```{r gcf-detail-table, cache=TRUE}
# join domains with GCF assignments first
domains_gcf <- domains |>
  mutate(bgc_id = paste0(sample_id, "__", contig_id)) |>
  left_join(gcf |> select(bgc_id, gcf_id), by = "bgc_id")

# dominant domains per GCF (top 5)
top_domains <- domains_gcf |>
  filter(gcf_id %in% top20$gcf_id) |>
  separate_rows(domain_array, sep = "\\|") |>
  filter(domain_array != "") |>
  count(gcf_id, domain_array, sort = TRUE) |>
  group_by(gcf_id) |>
  slice_max(n, n = 5) |>
  summarise(top_domains = paste(domain_array, collapse = "; "), .groups = "drop")

# mean BGC probability (deepBGC/GECCO only)
mean_prob <- gcf_meta |>
  filter(gcf_id %in% top20$gcf_id & !is.na(BGC_probability)) |>
  mutate(BGC_probability = as.numeric(BGC_probability)) |>
  group_by(gcf_id) |>
  summarise(mean_prob = round(mean(BGC_probability, na.rm = TRUE), 3),
            .groups = "drop")

# dominant backbone and product
gcf_char <- gcf_meta |>
  filter(gcf_id %in% top20$gcf_id) |>
  group_by(gcf_id) |>
  summarise(
    size     = n(),
    backbone = names(sort(table(backbone_class),     decreasing = TRUE))[1],
    product  = names(sort(table(product_harmonised), decreasing = TRUE))[1],
    .groups  = "drop"
  )

top20 |>
  select(gcf_id, n_cancer, n_noncancer, prev_cancer, prev_noncancer,
         odds_ratio, p_value, p_adj, sig) |>
  left_join(gcf_char,    by = "gcf_id") |>
  left_join(mean_prob,   by = "gcf_id") |>
  left_join(top_domains, by = "gcf_id") |>
  mutate(
    prev_cancer    = scales::percent(prev_cancer,    accuracy = 0.1),
    prev_noncancer = scales::percent(prev_noncancer, accuracy = 0.1),
    odds_ratio     = round(odds_ratio, 2),
    p_value        = formatC(p_value,  format = "e", digits = 2),
    p_adj          = formatC(p_adj,    format = "e", digits = 2)
  ) |>
  knitr::kable(
    col.names = c("GCF", "n Cancer", "n Non-cancer",
                  "Prev. Cancer", "Prev. Non-cancer",
                  "OR", "p", "p_adj (BH)", "Sig.",
                  "Size", "Backbone", "Product",
                  "Mean prob.", "Top domains")
  ) |>
  kable_styling(font_size = 11) |>
  scroll_box(width = "100%")
```

## Continent distribution

```{r continent}
meta |>
  count(Continent, group) |>
  ggplot(aes(x = Continent, y = n, fill = group)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  labs(x = NULL, y = "Genomes", fill = NULL,
       title = "Sample distribution by continent and clinical group") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Age distribution

```{r age}
meta |>
  filter(!is.na(Age)) |>
  mutate(Age = as.numeric(Age)) |>
  ggplot(aes(x = group, y = Age, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1.2) +
  scale_fill_manual(values = c("Cancer" = "#d7191c", "Non-cancer" = "#2c7bb6")) +
  labs(x = NULL, y = "Age (years)", fill = NULL,
       title = "Age distribution by clinical group") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```